Effects of pulsed hollow electron lens operation on the beam core in HL-LHC: \\First experimental studies and simulations.

### 2018-01-25 resex.tex
revised until 10th turn pulsing

\subsection{Excitation with transverse damper and filling scheme\label{sec:adt}}
\textcolor{red}{Daniel's part}
\begin{itemize}
	\item error on amplitude
	\item synchronization of H and V kicker
	\item random noise on amplitude - is there any?
	\item bandwidth - why does the damper not affect the nth turn pulsing????
	\item filling scheme
\end{itemize}

\subsection{Overview of simulations\label{sec:sim}}
For the preparation and interpretation of the experiments two different types of simulations were performed:
\begin{enumerate}
	\item tracking of a Gaussian distribution (Distribution Tracking) to obtain loss rates and emittance growth,
	\item Frequency map analysis (FMA)~\cite{fmalaskar} for the visualization of resonances.
\end{enumerate}
For both simulation types, the tracking code Lifetrac \cite{lifetrac} was used and the simulation parameters are summarized in Table~\ref{tab:sim_param}. Further details on the simulations can be found in \cite{md_sim_hel_res_ex_fitterer,md_sim_hel_res_ex_fitterer_update}.
\begin{table}[b]
	\caption{\label{tab:sim_param}%
		Summary of simulation parameters for the Distribution Tracking and FMA analysis. For details see \cite{md_sim_hel_res_ex_fitterer,md_sim_hel_res_ex_fitterer_update}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & Distribution Tracking & FMA \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{450 GeV} \\
			emittance & $3.5~\mu$m& $2.5~\mu$m\\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			particle & 6D Gaussian & equally spaced grid\\
			distribution &  distribution & in $x,y$ up to $10~\sigma$\\
			&  ($10^4$ particles) & with $\frac{\Delta p}{p_0}=0$\\
			turns tracked & $10^6$ turns & $10^4$ turns \\\hline
			optics & \multicolumn{2}{c}{2016 or 2017 injection optics,}\\
			& \multicolumn{2}{c}{$\beta^*=11$~m}\\
			machine  & standard errors & no errors \\
			imperfections &with $a_1=b_1=0$\footnote{Orbit errors are disabled due to different implementation of the $a_1,b_1$ errors in Lifetrac and MAD-X. $b_2$ errors are adjusted to yield an average (over 60 seeds) peak $\beta$-beat of 15\% as expected from optics measurements in the LHC.} &  \\
			octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}}=+19.6~\mathrm{A}$, $I_{\mathrm{MOD}}=-19.6~\mathrm{A}$}\\
			tune $(Q_x,Q_y)$ & \multicolumn{2}{c}{(64.28,59.31) for 2016,}\\
			& \multicolumn{2}{c}{(62.27,60.295) for 2017}\\
			chromaticity & \multicolumn{2}{c}{(+15,+15)}\\
			 $(Q'_x,Q'_y)$ & & \\\hline
			 transv. aperture & $5.7~\sigma$ & - \\
			 long. aperture & $10~\sigma$ & - \\
		\end{tabular}
	\end{ruledtabular}
\end{table}

\subsection{Simulation and experimental results\label{sec:simex}}
In the following we will summarize the simulation and experimental results. We will jump forth and back between both as the experimental results are best understood on the base of the simulations.

In summary, the most efficient pulsing patterns and the general tendency of the distribution changes could be successfully predicted in simulations (Sec.~\ref{sec:pattern}). Quantitative predictions of loss rates and emittance growth though failed as both observables are influenced also by the natural noise present in the LHC like ground motion and tune modulation. At injection this natural noise component is now known for the LHC. The observed effect is then the result of the interaction of the external excitation with the natural noise present.

In case of a resonant excitation, the beam distribution adjusts to a new equilibrium distribution. In the presence of only a resonant excitation, this distribution is stable and stays unchanged. If the resonant excitation interacts with the natural noise sources, the beam distribution still changes to a new equilibrium distribution but this change is also accompanied by a resonant excitation amplitude dependent emittance growth. Both effects were experimentally observed in case of the $10^{\mathrm{th}}$~turn pulsing (Sec.~\ref{sec:simex10}).

In case of an excitation with random uniform noise, the shape of the distribution does stay Gaussian. The interaction with the natural noise source then only manifests itself in an underestimation of loss rates and emittance growth (see Sec.~\ref{sec:simexran}). This effect was also studied in more detail in \cite{md1433_noise_top_energy,md_noise_bbLHC} for colliding beams at the LHC.

\subsubsection{Prediction of the most efficient pulsing pattern\label{sec:pattern}}
\begin{figure}[b]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_sigm.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit1.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:patternsim} Relative intensity (top left), bunch length (top right) and horizontal (bottom left) and vertical (bottom right) emittance obtained from simulation results (Distribution Tracking) for different pulsing patterns based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The excitation was applied in both planes with an amplitude of 96~nrad.}
\end{figure}
In order to predict the most efficient pulsing pattern, the loss rate and emittance growth from the Distribution Tracking are compared for the different pulsing patterns. The results for the 2016 optics and tune are shown in Fig.~\ref{fig:patternsim}. The largest losses are observed for $3^{\mathrm{rd}}$, $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing and the random excitation. As the bunch length decreases accordingly, the losses are mostly associated to off-momentum particles. Only for the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing as well as random excitation a considerable emittance growth is visible, where the random excitation has by far the strongest effect compared to any $k^{\mathrm{th}}$~turn pulsing pattern.

\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png}
	\end{minipage}
	\caption{\label{fig:patternfma} FMA analysis for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. The absence of a strong excitation of any resonance in case of $7^{\mathrm{th}}$~turn pulsing only in V and the strong excitation in case of pulsing only in H further confirms that indeed the $7Q_x$ resonance is driven. For $10^{\mathrm{th}}$~turn pulsing there is in contrast no significant difference between pulsing only in H, only in V or in H+V.}
\end{figure}
The sensitivity to the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing is also observed in absence of machine errors in which case the only source of non-linearities are the strong sextupoles and octupoles \cite{md_sim_hel_res_ex_fitterer}. The FMA analysis shown in Fig.~\ref{fig:patternfma} reveals that 
the $7Q_x$ resonance is excited in case of the $7^{\mathrm{th}}$~turn pulsing and the $10Q_x$ and $10Q_y$ resonance in case of the $10^{\mathrm{th}}$~turn pulsing. As octupoles can only drive even resonances, the source of the $7Q_x$ resonances are the sextupoles while the octupoles only generate the tune footprint. The other pulsing patterns do not exhibit any increase in losses or emittance growth without magnetic errors and their effect thus is based on the magnetic field errors assigned, explicitly also the seed chosen for the simulations.

\begin{figure}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from simulation results (Distribution Tracking) based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The excitation was applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here $10^4$~turns) minus the initial distribution (here $10^2$~turns), and the ratio as the final distribution divided by the initial distribution. The red line in the residual shows the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure}
For the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing the emittance also appears to start at an increased initial value and then stays almost constant. This behavior is due to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium distribution. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}. Besides an increased emittance, the new distribution also becomes non-Gaussian as can be seen from the residual in respect to the Gaussian distribution in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing (Fig.~\ref{fig:patternhist}). For comparison, also the distribution in both planes for a random excitation is shown This initial fast change of the beam distribution was also to the knowledge of the authors for the first time experimentally observed (Sec.~\ref{sec:simex10} and Sec.~\ref{sec:simex7}). The timescale of this fast adjustment is experimentally however much slower than predicted in simulations - on the timescale of minutes compared to seconds.

\subsubsection{10th turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing was tested in 2016 with an excitation in the vertical plane only. In summary, the following experimental results were obtained:
\begin{enumerate}
	\item increase of the loss rate with the excitation amplitude (Fig.~\ref{fig:10thexp}, \ref{fig:10thexpsim}),
	\item increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane (Fig.~\ref{fig:10thexp}).
	\item change of the beam distribution due to the resonant excitation (Fig.~\ref{fig:10thexp}, \ref{fig:10thexpprof})
\end{enumerate}
As loss rates, emittance growth and distribution stayed unchanged for the 4~reference bunches, the above observations can be truly associated with the effect of the resonant excitation.
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.49\linewidth]{bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and emittance (bottom) obtained during the 2016 experiments. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad, the excitation amplitude was linearly increased for each group of 4~bunches (see Sec.~\ref{sec:adt}). The different colored solid lines labeled with $n\Delta A$ indicate the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation as their envelope.}
\end{figure}

Besides the determination of the scaling of the loss rate and emittance growth with the excitation amplitude as required for the specification of the hollow electron lens, the most novel observation is the direct measurement of the change of the beam distribution, here with the Beam Synchrotron Radiation Telescope (BSRT) \cite{bsrtprofinj}. As an example for the change in distribution, the vertical profile of a reference bunch and one bunch experiencing the maximum excitation are compared in Fig.~\ref{fig:10thexpprof}. The distribution of the excited bunch clearly changes and becomes non-Gaussian, while the reference bunch stays almost unchanged.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_50.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_1300.png}
	\end{minipage}
	\caption{\label{fig:10thexpprof} Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2016 experiments. The profiles are taken at the end of the $10^{\mathrm{th}}$~turn pulsing in V. The black line in the plot of the residual is the current profile minus the initial profile and is a measure for the overall change of the distribution. The red line is the current profile minus its Gaussian fit, which gives and indication for the deviation of the distribution from a Gaussian distribution. For details on the analysis see \cite{bsrtprofinj}. The distribution of the reference bunch (left), here slot 50, stays unchanged, while the bunch experiencing the maximum excitation (right), here slot 1300, clearly shows a change to a non-Gaussian distribution.}
\end{figure}

Comparing these experimental results with the simulations including only the $10^{\mathrm{th}}$~turn pulsing (solid line, Fig.~\ref{fig:10thsim}), the simulations predict qualitatively correctly an excitation amplitude dependent loss rate, but fail to predict the excitation amplitude dependent, continuous emittance growth seen experimentally.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2_rel.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:10thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$. The solid line indicates an excitation with only $10^{\mathrm{th}}$~turn pulsing in V and the dotted line with $10^{\mathrm{th}}$~turn pulsing in V plus a random dipole noise component in H+V of 6~nrad.}
\end{figure}
Instead, the simulations yield a constant emittance after an initial increase during the first $10^4$~turns or first second. So obviously there is an essential ingredient missing. As it turns out, the constant emittance can be changed to an excitation amplitude dependent emittance growth by adding a random dipole noise component with uniform distribution (dotted line in Fig.~\ref{fig:10thsim}). As a first estimate the value obtained in \cite{md1433_noise_top_energy,md_noise_bbLHC} for 6.5~TeV was used and scaled by the magnetic rigidity to the injection energy of 450~GeV, as no noise estimate is available at injection. This yields a maximum kick amplitude at the transverse damper of approximately
\begin{equation}
\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad},
\end{equation}
Adding this random noise component, the simulations then yield correctly an excitation amplitude dependent emittance growth in the vertical plane only after a first fast adjustment of the beam distribution visible as an increased initial value.

>>>>

\subsection{Excitation with transverse damper and filling scheme\label{sec:adt}}
The primary function of the transverse feedback system in LHC, also referenced to as ADT, is to provide injection oscillation damping and actively damp the coupled bunch instabilities driven by the machine impedance. The main building blocks of the ADT system are stipline pickups at the position Q7 and Q9 at Point~4 of the LHC, connected to the beam position measurement modules at the surface, the digital signal processing modules (mDSPU) and a set of tetrode power amplifiers feeding electrostatic kickers in the RF zone of the LHC \cite{adt_sum_2008,adt_sum_2011}.

Though originally designed to damp the undesired transverse activities only, the damper is being routinely used for sophisticated beam excitations as well (e.g. Abort gap and injection gap cleaning, single bunch excitation for tune and linear coupling measurements, machine development etc.) thanks to the state of the art digital signal processing hardware.

The transverse feedback is in general active during the whole LHC cycle, therefore it is of very high importance, that the noise introduced by the system does not contribute to any emittance growth. The typical machine cycle requires a short damping time of ~10-20 turns (high feedback gain) for injection oscillation damping, which is increased during the ramp and stable beams to ~50-100 turns (lower feedback gain). It has been demonstrated that noise introduced by the transverse feedback system itself is sufficiently low to introduce any measurable emittance growth \cite{adt_noise_emit_2017}. All excitation signals referenced to in this paper, are digitally synthesized in the damper's digital signal processing units and therefore can be considered as ``noise-free''.

The resonant excitation experiment in 2017 involved simultaneous measurements on three groups of 72 bunches with a dedicated excitation pattern and transverse feedback configuration on each sub-group of 6 bunches. In 2016 a similar configuration was used, but with only 48 bunches in total. Both schemes are illustrated in Fig.~\ref{fig:fill}.
\begin{figure*}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		2016 experiment
		\includegraphics[width=1.0\linewidth]{bunchfilling.png}	
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
	\centering
	2017 experiment
	\includegraphics[width=1.0\linewidth]{bunchfilling.png}	
	\end{minipage}
	\caption{\label{fig:fill} Filling and excitation scheme for LHC experiments in 2016 (top) and 2017 (bottom).}
\end{figure*}

The excitation in the horizontal and the vertical plane is generated by a different set of signal processing hardware, however during this experiment the signals were synchronized at a turn level. This means the bunches of the third group (of 72 bunches) experienced the kick in both horizontal and vertical plane during the same turn.

The ADT kicker deflection angle is calculated from the kicker geometry and the excitation voltage on the deflection plates. The excitation voltage used in the experiments is a complicated waveform, generated by a very complex hardware (digital signal processor, chain of transmission lines, low and high power amplifiers). The peak excitation voltage (and subsequently the kick) seen by the beam can be only indirectly measured using the probes on each kicker. The measured values are therefore believed to be known with an error margin of 10-15\% for the experiments in 2017 and 50\% in 2016. The error for 2016 is much larger as no calibration of the ADT was performed beforehand, which was done in 2017. As an example Fig.~\ref{fig:fill_meas} shows the oscillation amplitude of each bunch in both H and V planes captured by the LHC real time transverse activity monitor. The measured activity is well in line with the expected excitation pattern.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{bunchfilling_measured.png}	
	\caption{\label{fig:fill_meas} Bunch by bunch excitation detected by the real time transverse activity monitor during the 2017 experiment (8th turn pulsing}
\end{figure}

\subsection{Overview of simulations\label{sec:sim}}
For the preparation and interpretation of the experiments two different types of simulations were performed:
\begin{table}[b]
	\caption{\label{tab:sim_param}%
		Summary of simulation parameters for the Distribution Tracking and FMA analysis. For details see \cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & Distribution Tracking & FMA \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{450 GeV} \\
			emittance & $3.5~\mu$m& $2.5~\mu$m\\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			particle & 6D Gaussian & equally spaced grid\\
			distribution &  distribution & in $x,y$ up to $10~\sigma$\\
			&  ($10^4$ particles) & with $\frac{\Delta p}{p_0}=0$\\
			turns tracked & $10^6$ turns & $10^4$ turns \\\hline
			optics & \multicolumn{2}{c}{2016 or 2017 injection optics,}\\
			& \multicolumn{2}{c}{$\beta^*=11$~m}\\
			machine  & standard errors & no errors \\
			imperfections &with $a_1=b_1=0$\footnote{Orbit errors are disabled due to different implementation of the $a_1,b_1$ errors in Lifetrac and MAD-X. $b_2$ errors are adjusted to yield an average (over 60 seeds) peak $\beta$-beat of 15\% as expected from optics measurements in the LHC.} &  \\
			octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}}=+19.6~\mathrm{A}$, $I_{\mathrm{MOD}}=-19.6~\mathrm{A}$}\\
			tune $(Q_x,Q_y)$ & \multicolumn{2}{c}{(64.28,59.31) for 2016,}\\
			& \multicolumn{2}{c}{(62.27,60.295) for 2017}\\
			chromaticity & \multicolumn{2}{c}{(+15,+15)}\\
			 $(Q'_x,Q'_y)$ & & \\\hline
			 transv. aperture & $5.7~\sigma$ & - \\
			 long. aperture & $10~\sigma$ & - \\
		\end{tabular}
	\end{ruledtabular}
\end{table}
\begin{enumerate}
	\item tracking of a Gaussian distribution (Distribution Tracking) to obtain loss rates and emittance growth,
	\item Frequency map analysis (FMA)~\cite{fmalaskar} for the visualization of resonances.
\end{enumerate}
For both simulation types, the tracking code Lifetrac \cite{lifetrac} was used. The simulation parameters are summarized in Table~\ref{tab:sim_param}. Further details on the simulations can be found in \cite{md_sim_hel_res_ex_fitterer,resexmd2017}.

\subsection{Simulation and experimental results\label{sec:simex}}
In the following we will summarize the simulation and experimental results. We will jump forth and back between both as the experimental results are best understood on the base of the simulations. In summary, the most efficient pulsing patterns could be successfully predicted in simulations (Sec.~\ref{sec:pattern}). Quantitative predictions of loss rates and emittance growth proved however challenging as both observables are influenced in addition by the natural noise present in the LHC originating for example from ground motion and magnetic field ripple. The observed effect is then the result of the interaction of the external excitation with the natural noise. At injection, this natural noise component is not known for the LHC. Beside the natural noise, also collective effects like intra-beam scattering and electron cloud influence the time evolution of emittance and losses. Both are not taken into account in the simulations.

The general dynamics of a resonant excitation seen in simulations is a fast adjustment to a new equilibrium distribution. In the presence of only a resonant excitation, this distribution is stable with a constant emittance as shown for example in Figs.~\ref{fig:patternsim} and \ref{fig:10thsim}. If the resonant excitation interacts with the natural noise sources, the beam distribution still changes to a new equilibrium distribution but this change is also accompanied by a resonant excitation amplitude dependent emittance growth as seen for example in Figs.~\ref{fig:10thsim},\ref{fig:7thsim} and \ref{fig:8thsim}, and also observed experimentally, see for example Figs.~\ref{fig:10thexp} and ~\ref{fig:7thexp}. An excitation with random uniform noise on the other hand results in constant emittance growth without any initial adjustment phase. The interaction with the natural noise source then only manifests itself in an underestimation of loss rates and emittance growth. This is visible in simulations, see Fig.~\ref{fig:ransim} as well as experimentally, see Fig.~\ref{fig:ranexp}. The effect of a random excitation was also studied in more detail in \cite{md1433_noise_top_energy,md_noise_bbLHC} for colliding beams at the LHC.

\subsubsection{Prediction of the most efficient pulsing pattern\label{sec:pattern}}
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_intensity.png}
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_sigm.png}
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit1.png}
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit2.png}
	\caption{\label{fig:patternsim} Relative intensity (top left), bunch length (top right) and horizontal (bottom left) and vertical (bottom right) emittance obtained from simulation results (Distribution Tracking) for different pulsing patterns based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The resonant excitation is applied in both planes with an amplitude of 96~nrad. No random noise component is added.}
The most efficient pulsing pattern can be predicted by the comparison of the loss rate and emittance growth from the Distribution Tracking for the different pulsing patterns. As an example, the simulation results for the 2016 experiment are shown in Fig.~\ref{fig:patternsim} featuring a clear dependence of both parameters on the pulsing pattern. The largest losses are observed for $3^{\mathrm{rd}}$, $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing and a uniform random excitation. As the bunch length decreases in accordance with the losses, the losses can be mostly associated to off-momentum particles. Only for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing as well as for a random excitation considerable emittance growth is visible. Compared to any $k^{\mathrm{th}}$~turn pulsing pattern, the random excitation has by far the strongest effect. The sensitivity to $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing is also observed in absence of machine errors in which case the only source of non-linearities are the strong sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer}. This suggests that latter are the source of non-linearity responsible for this sensitivity. The driven resonances are revealed by the FMA analysis shown in Fig.~\ref{fig:patternfma}. The $7Q_x$ resonance is excited in case of the $7^{\mathrm{th}}$~turn pulsing and the $10Q_x$ and $10Q_y$ resonance in case of the $10^{\mathrm{th}}$~turn pulsing for an excitation in both planes. As octupoles can only drive even resonances, the source of the $7Q_x$ resonances are the sextupoles while the octupoles only generate the tune footprint. The other pulsing patterns do not exhibit any increase in losses or emittance growth without magnetic errors and their effect thus is based on the magnetic field errors assigned, explicitly also the seed chosen for the simulation.
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png}
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
		$7^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png}
		$7^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png}
	\caption{\label{fig:patternfma} FMA analysis for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. The absence of a strong excitation of any resonance in case of $7^{\mathrm{th}}$~turn pulsing only in V and the strong excitation in case of pulsing only in H further confirms that the $7Q_x$ resonance is the main resonance driven. For $10^{\mathrm{th}}$~turn pulsing there is in contrast no significant difference between pulsing only in H, only in V or in H+V (see Appendix~\ref{sec:fma:10}, Fig.~\ref{fig:fma:10}).}

\begin{figure}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn pulsing (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from Distribution Tracking simulations based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ and standard errors. The excitation is applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here $10^4$~turns) minus the initial distribution (here $10^2$~turns), and the ratio as the final distribution divided by the initial distribution. The red line in the plot of the residual is the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure}

As already addressed earlier, the emittance in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing starts at an increased initial value and then stays almost constant during the entire simulation time. This behavior is typical for any $n^{\mathrm{th}}$ turn pulsing and can be ascribed to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium distribution. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}, and also any other simulations over $10^6$ turns presented in this paper. Besides an increased emittance, the new distribution also becomes non-Gaussian. This can be seen by means of the residual in respect to the Gaussian distribution illustrated for example for the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing in Fig.~\ref{fig:patternhist}. For the purpose of comparison, also the distribution in both planes for a random excitation is shown. This initial fast change of the beam distribution is not only an artifact of the simulation, but has been, to the knowledge of the authors, also for the first time experimentally observed as will be presented later in this paper. The timescale of this fast adjustment is in experiments however much slower than predicted in simulations - on the timescale of minutes compared to seconds.

\subsubsection{10th turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. In summary, the following experimental results were obtained:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude as shown in Figs.~\ref{fig:10thexp} and \ref{fig:10thexpsim}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane shown in Fig.~\ref{fig:10thexp}.
	\item Change of the beam distribution due to the resonant excitation illustrated in Fig.~\ref{fig:10thexp} and \ref{fig:10thexpprof}.
	\item As loss rates, emittance growth and distribution stayed unchanged for the $4$~reference bunches, the above observations can be truly associated with the effect of the resonant excitation.	
\end{enumerate}
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.49\linewidth]{bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and emittance (bottom) obtained during the 2016 experiments. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad, the excitation amplitude was linearly increased for each group of 4~bunches (see Sec.~\ref{sec:adt}). The different colored solid lines labeled with $n\Delta A$ indicate the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation as their envelope.}
\end{figure}
